---
title: "Part 4: Multiple Regression and Logistic Regression"
author: "Sam Tyner"
date: "TBD"
output: 
  beamer_presentation:
    theme: "AnnArbor"
    colortheme: "dolphin"
    includes:
      in_header: header.tex
      before_body: slide_prefix.tex
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, fig.align = 'center', fig.width = 6, cache=T)
```

```{r loadpkgs}
library(tidyverse)
```

## Textbook

These slides are based on the book *OpenIntro Statistics* by David Diez, Christopher Barr, and Mine Çetinkaya-Rundel

The book can be downloaded from [https://www.openintro.org/stat/textbook.php](https://www.openintro.org/stat/textbook.php)

Part 4 Corresponds to Chapter 8 of the text. Sections 3.1-3.3 correspond to sections 3.1, 3.4, 3.5.2 of the text.

## Outline 

- Multiple Regression (4.1)
- Model Selection (4.2)
- Logistic Regression (4.3)

# Section 4.1: Multiple Regression

# Section 4.2: Model Selection

# Section 4.3: Logistic Regression