---
title: "Reading Files"
author: "Sam Tyner, Haley Jeppson"
date: "2018/06/06"
output:
  xaringan::moon_reader:
    css: ["default", "csafe.css", "csafe-fonts.css"]
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message= FALSE, warning = FALSE)
```

class: primary
# Data in Excel

- Formats xls and csv - what's the difference?

- File extensions xls and xlsx are proprietary Excel formats, binary files

- csv is an extension for Comma Separated Value files. They are text files - directly readable.

- Example: Glass element analysis data

---
class: primary
# Glass data

Visit [this link](https://raw.githubusercontent.com/CSAFE-ISU/REU-materials-2018/master/slides/rstats/data/GK_clean_results.csv) to get the data. 

```{r message=FALSE, warning=FALSE}
library(readr)
glass <- read_csv("https://raw.githubusercontent.com/CSAFE-ISU/REU-materials-2018/master/slides/rstats/data/GK_clean_results.csv")
head(glass)
```


---
class: primary
# `read_csv` vs. `read_*`

`read_csv` is just one way to read a file using the `readr` package: 

- `read_delim`: the most generic function. Use the `delim` argument to read a file with any type of delimiter 
- `read_tsv`: read tab separated files 
- `read_lines`: read a file into a vector that has one element per line of the file
- `read_file`: read a file into a single character element
- `read_table`: read a file separated by space

---
class: primary
# Now we can make a plot!

```{r, warning=FALSE, message=FALSE, fig.width=7, fig.height=4}
library(ggplot2)
ggplot(glass, 
       aes(x = Li7, y = Na23)) + 
  geom_point()
```

---
class: primary
# Read Excel Data

Download the boxoffice.xlsx file from [this link](https://github.com/CSAFE-ISU/REU-materials-2018/blob/master/slides/rstats/data/boxoffice.xlsx) and put it in your "LearningR" folder. 

Read it in and save it as `boxofc`

```{r, eval=FALSE}
library(readxl)
 
boxofc <- read_excel("boxoffice.xlsx")

head(boxofc)
```
```{r, echo=FALSE}
library(readxl)
 
boxofc <- read_excel("data/boxoffice.xlsx")

head(boxofc)
```

---
class: primary
# `foreign` Package

- Other file formats can be read using the functions from package `foreign`

- SPSS: `read.spss`

- SAS: `read.xport`

- Minitab: `read.mtp`

- Systat: `read.systat`

---
class: primary
# Your Turn 3.1 

The NHANES (National Health and Nutrition Survey) publishes data in the SAS xport format:

https://wwwn.cdc.gov/Nchs/Nhanes/continuousnhanes/default.aspx?BeginYear=2013

1. Scroll to the bottom, choose one of the datasets (Demographics, Examination, etc.). Download the Data file (XPT)

2. Use `read.xport()` to load the file into R

3. Briefly examine the dataset you've imported (use `head` or `tail`, etc)

---
class: primary
# Other Data Types 

What can you think of? 

---
class: primary
# Google Sheets 

```{r eval = FALSE}
install.packages("googlesheets")
library(googlesheets)
gap <- gs_key("1D5FK6z7aGukdGRAJQoyuPc2QV5zSJ4oN7oslp_u7YEQ")
gs_browse(gap)
africa <- gs_read(gap, "Africa")
americas <- gs_read(gap, "Americas")
```
```{r echo = FALSE}
library(readxl)
africa <- read_xlsx("data/Gapminder.xlsx", sheet = "Africa")
head(africa)
```

---
class: primary
# Text files 

Read the data [here](https://raw.githubusercontent.com/CSAFE-ISU/REU-materials-2018/master/slides/rstats/data/poem.txt)

```{r}
poem <- read_lines("https://raw.githubusercontent.com/CSAFE-ISU/REU-materials-2018/master/slides/rstats/data/poem.txt")
poem
```

---
class: primary
# Your Turn 3.2

Read in the rest of the Google Sheets from the Gapminder data set and combine them into one data set with the `bind_rows()` function. Save it as `gapdata`.



